---
title: "Northeast '115' 4000-footers"
subtitle: "TidyTuesday 2026-01-11"
date: "2026-01-11"
author: "Aaron Wright [`aawva@mit.edu`]"
categories: [r, ggplot2, tidyverse, data-viz]
image: "plot.svg"
---

<!-- THEMATICS AND RENDERING -->

::: cell

```{=html}
<style>
/* inline code */
code {
  background-color: #f5f5f5;
  color: #859901; /* solarized green */
  border-radius: 4px;
  padding: 2px 4px;
  font-family: monospace;
  font-size: 0.95em;
}

/* list item spacing for inline code */
li code {
  margin-left: 2px;
}

/* link styling */
a {
  color: #2AA198; /* solarized cyan */
  text-decoration: none;
}
a:hover {
  text-decoration: underline;
}
</style>
```
:::

<!-- CONFIGURATION -->

## Overview
---
**Research Question:** {{RESEARCH_QUESTION}}

Do climbers favor tall peaks, prominent peaks, or just accessible ones?
How do elevation distributions differ by range within the same state?

This week's TidyTuesday challenge focused on [DATASET_DESCRIPTION]. I explored [KEY_INSIGHTS] using r to [VISUALIZATION_APPROACH], asking [RESEARCH_QUESTION]

> **Key Libraries**: ggplot2, dplyr, tidyr <br>

<!-- ENVIRONMENT & DATA -->

## Dataset
---
The dataset for this week included [DATASET_FEATURES].

**Elevation vs. prominence:**
Elevation refers to how high a peak rests above sea level (absolute measure), while prominence denotes how much it sticks up compared to the surrounding terrain (relative measure). E.g., elevation would signal a building’s total height, and prominence would capture how much taller it is than the buildings surrounding it.

```
ne115 dataset
- peak — official summit name
- rank_ne — AMC ordering of peaks by elevation (1 = highest)
- rank_state - relative within-state ranking
- rank_range - relative within-region ranking
- state — US state in which the peak resides
- range — mountain range containing the peak (e.g., Adirondacks, Whites, Greens)
- ascents — number of recorded ascents (typically from list registries, guidebook logs, or community reports)
- elevation_ft — summit elevation (feet above sea level)
- elevation_m — summit elevation (meters above sea level)
- prominence_ft — vertical rise of the peak, relative to the lowest contour connecting it to a higher peak (feet)
- prominence_m — prominence (meters)
- key_col_ft - lowest point of necessary descent from a peak before one could climb to a higher one (defining prominence) (feet)
- key_col_m - key col (meters)
- prominence_ratio — the fraction of a peak’s elevation represented by its prominence, indicating how independently it rises above surrounding terrain (values near 1 = highly independent; near 0 = less independent)
```

> **Source:** [Northeast "115" 4000-footers](https://www.peakbagger.com/list.aspx?lid=511&pt=prom&hu=0&w=0&u=ft) <br>
> **Inspiration:** [Ridgeline plot with inside plot and annotations](https://r-graph-gallery.com/web-ridgeline-plot-with-inside-plot-and-annotations.html) <br>

### Environment

```{r}
#| label: environment
#| message: false
#| warning: false

# Load required libraries
library(tidytuesdayR)
library(tidyverse)
library(here)
library(rvest)
library(janitor)
library(camcorder) # not supported in Positron
library(ggridges)
library(lalaland) # La La Land color palettes for viz
source("scripts/viz_aesthetics.R") # personal aesthetics
source("scripts/mcc.R") # "makeshift camcorder (mcc)" - custom {camcorder} imitation for Positron
```

### Import Peaks Data

```{r}
#| label: peaks data
#| message: false
#| warning: false

# save the html to repo
# https://www.peakbagger.com/list.aspx?lid=511&pt=prom&hu=0&w=0&u=ft (english units)
# https://www.peakbagger.com/list.aspx?lid=511&pt=prom&hu=0&w=0&u=m (metric units)

# point to html for scraping peaks table
html_path_ft <- here("challenges", "2026", "2026-01-11_01_Northeast_115_4000-footers", "data", "peakbagger_ne115_4000_ft.html")
html_path_m <- here("challenges", "2026", "2026-01-11_01_Northeast_115_4000-footers", "data", "peakbagger_ne115_4000_m.html")

# helper function to extract mountains with both metrics from saved html files
get_peaks_df <- function(html_path) {

  # read saved html file
  page <- read_html(html_path)
  
  # extract all table elements
  tables <- page %>% html_elements("table")
  
  # convert tables to dfs
  tbl_list <- map(tables, html_table)
  
  # select target table by index
  peaks_raw <- tbl_list[[5]] # index 5 points to 5th table element on page, which holds the peaks data
  
  # clean column names for tidy workflows
  peaks <- peaks_raw %>% clean_names()
  
  # return tibble
  peaks
}

# extract peaks dfs
peaks_ft <- get_peaks_df(html_path = here("challenges", "2026", "2026-01-11_01_Northeast_115_4000-footers", "data", "peakbagger_ne115_4000_ft.html"))
peaks_m <- get_peaks_df(html_path = here("challenges", "2026", "2026-01-11_01_Northeast_115_4000-footers", "data", "peakbagger_ne115_4000_m.html"))
```

<!-- WRANGLING & ANALYSIS -->

## Analysis
---
### Cleaning

```{r}
#| label: cleaning
#| echo: true
#| message: false
#| warning: false

# clean and prepare the data
ne115 <- peaks_ft %>%
  left_join(peaks_m, by = c("rank", "peak", "state")) %>% # join english and metric dfs
  mutate(
    range = coalesce(range_level_5.x, range_level_5.y), # same in both dfs
    ascents = coalesce(ascents.x, ascents.y), # same in both dfs
    key_col_ft = elev_ft - prom_ft,
    key_col_m = elev_m - prom_m,
    prominence_ratio = prom_ft / elev_ft,   # % of elevation that is "independent"
    ) %>%
  group_by(state) %>%
  mutate(rank_state = min_rank(desc(elev_ft))) %>% # in-state rank
  ungroup() %>%
  group_by(range) %>%
  mutate(rank_range = min_rank(desc(elev_ft))) %>% # in-range rank
  ungroup() %>%
  rename(
    rank_ne = rank, # rank by elevation across ne
    elevation_ft = elev_ft, # informativity
    elevation_m = elev_m,
    prominence_ft = prom_ft,
    prominence_m = prom_m
  ) %>%
  select(
    peak, # reordering more intuitively :)
    rank_ne,
    rank_state,
    rank_range,
    state,
    range,
    ascents,
    elevation_ft,
    elevation_m,
    prominence_ft,
    prominence_m,
    key_col_ft,
    key_col_m,
    prominence_ratio
  )

# save ne115 df
write_csv(ne115, here("challenges", "2026", "2026-01-11_01_Northeast_115_4000-footers", "data","ne115_4000s.csv"))
```

## Visualization
---

```{r}
#| label: main visualization
#| fig-width: 12
#| fig-height: 8
#| warning: false

source("scripts/mcc.R")


# initialize makeshift camcorder
init_mcc(
  dir = here("challenges","2026","2026-01-11_01_Northeast_115_4000-footers","viz","gif","frames"),
  prefix = "ne115",
  width = 14,
  height = 12,
  dpi = 300
)

# create the main visualization

( # open parens for mcc

ggplot(ne115, aes(x = elevation_ft, y = state)) +
  geom_density_ridges() +

) %>% # close parens for mcc

  # resize_frame(width = 14, height = 12) %>%
  save_frame() # save plot with makeshift camcorder

# stop makeshift camcorder to stitch plotting gif
stop_mcc(gif_name = "ne115.gif")
```

<!-- VISUALIZATION & COMMENTARY -->

## Technical Summary
---
- **Data Source**: [Characteristics of Adirondack Hiking Trails (Key2STATS)](https://www.key2stats.com/data-set/view/835)
- **Analysis Date**: 2026-01-11
- **Tools Used**: R, ggplot2, tidyverse
- **Key Libraries**: ggplot2, dplyr, tidyr

## Vizualization
---
![]({IMAGE_FILE})

## Future Directions
---
- [POTENTIAL_IMPROVEMENTS]
- [FUTURE_ANALYSIS_IDEAS]
- [DATA_QUALITY_NOTES]

<!-- CITATIONS -->

## TidyTuesday
---
> **Official Tidy Tuesday Repository**: [tidytuesday](https://github.com/rfordatascience/tidytuesday) <br>
> **{tidytuesdayR} Package Repository**: [tidytuesdayR](https://github.com/dslc-io/tidytuesdayR) <br>
> **CRAN Package**: [CRAN: Package tidytuesdayR](https://CRAN.R-project.org/package=tidytuesdayR) <br>
> **References** @tidytuesday, @tidytuesdayR